---
title: "Introduction to Simple Linear Regression"
author: "Amy Fishencord"
date: "2024-10-25"
output:
  ioslides_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(plotly)
```

<!--
The following two parts are responsible for the logo and the bar around the logo.
-->

<style>
.gdbar img {
 width: 150px !important;
 height: 80px !important;
 margin: 8px 8px;
}

.gdbar {
 width: 200px !important;
 height: 100px !important;
}

</style>

<!--
The following parts are responsible for the text. Actually, it specifies certain features
of the text. p: paragraph, div: division (it's similar to the paragraph), h1: heading at 
level 1 (the largest title or the largest heading), h2 is a bit smaller heading (h2 is the
title of each slide)...
Below we are specifying that all of these html tags will have color of the text black
and font family: Modern Computer Roman. (This is the default Latex font). 
-->

<style type="text/css">
body,p,div,h1,h2,h3,h4,h5 {
  color: black;
  font-family: Modern Computer Roman;
}

h2{
color: deeppink

}


</style>

## Linear Regression
Linear regression is a statistical model that estimates the linear relationship between a scalar response and one or more explanatory variables.

Scientists in many fields, including biology and the behavioral, environmental, and social sciences, use linear regression to conduct preliminary data analysis and predict future trends.

In linear regression, the independent variable is the variable that is used to predict the value of another variable, while the dependent variable is the variable that is being predicted.

## Mathematical Representation
Equation of Simple Linear Regression:
$$
y = {\beta}_0 + {\beta}_1x
$$
where \(y\) is the dependent variable, 
\({\beta}_0\) is the intercept (when y = 0), 
\({\beta}_1\) is the slope of the regression line and 
\(x\) is the independent variable

Expected Value of \(y\):
$$
E(y) = {\beta}_0 + {\beta}_1x
$$
where,
\(E(y)\) is the expected value of \(y\),
\({\beta}_0\) is the intercept,
\({\beta}_1\) is the slope,
and \(x\) is the independent variable

## Data Set Used
We need to work with a data set to further explain linear regression. We will now load in the mtcars dataset.
```{r, echo = TRUE}
data("mtcars")
head(mtcars)
```

## Weight vs Mpg
```{r, fig.width=6, fig.height=4, fig.align='center', echo = TRUE}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(x = "Weight (in 1000's of lbs)", y = "Miles Per Gallon")
```

## Linear Model of Weight vs MGP
```{r, fig.width=6, fig.height=4, fig.align='center'}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "deeppink") +
labs(x = "Weight (in 1000's of lbs)", y = "Miles Per Gallon")
```
We can observe there is a negative regression line. This means that for every unit increase in the independent variable (like weight), the dependent variable (MPG) decreases.

## Linear Model Summary
```{r, echo = TRUE}
model <- lm(mpg ~ wt, data = mtcars)
wt_coefficient <- summary(model)$coefficients["wt", "Estimate"]
wt_coefficient
```
The coefficient for wt is -5.3445, meaning that for every increase of 1 unit in weight (1000 lbs), the mpg decreases by approximately 5.34. This indicates a negative relationship between weight and miles per gallon: as weight increases, fuel efficiency decreases.

- The coefficient for wt (\({\beta}_1\)) is the slope in the equation
\(E(y)\) = \({\beta}_0\) + \({\beta}_1x\)
- The variable wt itself represent \(x\) in the equation.

## Horsepower vs MPG Linear Model

<style>
.small-font {
  font-size: 0.8em; /* Adjust the size as needed */
}
</style>

<div class="small-font">

```{r, fig.width=4.5, fig.height=3.5, fig.align='center'}
ggplot(mtcars, aes(hp, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "pink") +
  labs(x = "Horsepower", y = "Miles Per Gallon")

model2 <- lm(mpg ~ hp, data = mtcars)
hp_coeff <- summary(model2)$coefficients["hp", "Estimate"]
hp_coeff
```
We can observe the linear relationship of horsepower vs miles per gallon having a negative relationship. The coefficient -0.06 represents the expected change in miles per gallon for each one-unit increase in horsepower. 

## Mathematical Representation
$$
E(y) = {\beta}_0 + {\beta}_1x
$$
From the previous graph, we are calculating the expected value (\(E(y)\)) of miles per gallon based on horsepower.

- \({\beta_0}\) is the intercept of the regression line, representing the expected MPG when horsepower is zero. While in practical terms, a horsepower of zero may not make sense, it provides a baseline for understanding the relationship.

- \({\beta}_1\) is the slope of the regression line, indicating the expected negative change of 0.06 in MPG for each one-unit increase in horsepower.

## Mathematical Representation Example 
For example, if we assume \({\beta_0}\) = 30 and \({\beta}_1\) = -0.06 based from our earlier analysis, we can say:

- When horsepower is zero, the model predicts that the MPG would be 30.

- For every additional horsepower, we expect the MPG to decrease by 0.06, suggesting that higher horsepower vehicles are generally less fuel-efficient.

By examining these coefficients and their meanings, we can better understand the relationship between vehicle characteristics and their performance in terms of fuel consumption.

## Behind the scenes 

The code the use to produce the graph and horsepower coefficient is shown below. 
```{r, echo = TRUE, fig.show = "hide", eval = FALSE}
ggplot(mtcars, aes(hp, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
              color = "pink") +
  labs(x = "Horsepower", y = "Miles Per Gallon")

model2 <- lm(mpg ~ hp, data = mtcars)
hp_coeff <- summary(model2)$coefficients["hp", "Estimate"]
hp_coeff
```

## 3D Scatter Plot
```{r,fig.width=7, fig.height=5.5, echo = FALSE}
plot_ly(data = mtcars, x = ~hp, y = ~wt, z = ~mpg,
        type = "scatter3d", mode = "markers", 
        marker = list(size = 5, color = "deeppink", opacity = 0.9)) %>%
  layout(scene = list(
      xaxis = list(title = 'Horsepower'),
      yaxis = list(title = "Weight (in 1000's of lbs)"),
      zaxis = list(title = 'Miles Per Gallon')),
    title = "3D Scatter Plot of MPG, Horsepower, and Weight"
)
```

## Behind the Scenes
The code used to produce the 3D scatter plot in the previous slide is provided below.
```{r, echo = TRUE, fig.show = "hide", eval = FALSE}
plot_ly(data = mtcars, x = ~hp, y = ~wt, z = ~mpg,
        type = "scatter3d", mode = "markers", 
        marker = list(size = 5, color = "deeppink", opacity = 0.9)) %>%
  layout(scene = list(
      xaxis = list(title = 'Horsepower'),
      yaxis = list(title = "Weight (in 1000's of lbs)"),
      zaxis = list(title = 'Miles Per Gallon')),
    title = "3D Scatter Plot of MPG, Horsepower, and Weight"
)
```

## Conclusion

<div class="small-font">

Key Takeaways:

- Linear regression models demonstrate the negative relationship between horsepower MPG as well as weight and MPG.
- The coefficients provide valuable insights into how changes in independent variables (horsepower and weight) affect fuel efficiency.

Implications: 

- Understanding these relationships can inform consumer choices when purchasing vehicles, emphasizing the trade-offs between performance (horsepower) and fuel efficiency.

Further Research:

- Further exploration of other factors influencing MPG, such as engine type and vehicle design, could enhance our understanding of fuel efficiency.

</div>
